# Financial Econometrics -- Problem Set 1

Daniel Campos, Joshua Castillo & Cherryl Chico. January 23, 2026.

## GDPUSA

First of all, let us load the data.

```{r}
gdpusa <- ts(read.table("GDPUSA.dat"), start = 1947, freq = 4)
gdpusa
```

We plot the time series to get a sense of its structure.

```{r}
plot(gdpusa, xlab = "Year", ylab = "US GDP (billions of USD)")
```

An overall upward trend in US GDP is clearly visible.

### Is the variance constant?

*Note: Grouping data every two years caused issues with some plots, so we opted to keep working with the time series as it is.*

We divide the time series in three windows and plot each one in order to check whether the variance increases over time.

```{r}
win1 <- window(gdpusa, start = 1947, end = 1972)
win2 <- window(gdpusa, start = 1972, end = 1997)
win3 <- window(gdpusa, start = 1997, end = 2019)

par(mfrow = c(1, 3))
plot(win1, ylim = c(min(gdpusa), max(gdpusa)))
plot(win2, ylim = c(min(gdpusa), max(gdpusa)))
plot(win3, ylim = c(min(gdpusa), max(gdpusa)))
```

Although it appears to be the case, we also compute the boxplot for each window to verify our suspicion.

```{r}
par(mfrow = c(1, 3))
boxplot(win1, ylim = c(min(gdpusa), max(gdpusa)))
boxplot(win2, ylim = c(min(gdpusa), max(gdpusa)))
boxplot(win3, ylim = c(min(gdpusa), max(gdpusa)))
```

It is evident that the variance is non-constant, increasing with the level of the time series. A logarithmic transformation is applied to fix this.

```{r}
lngdp <- log(gdpusa)
plot(lngdp)
```

The previous plots are drawn again to confirm that the transformation has indeed stabilized the variance.

```{r}
lnwin1 <- window(lngdp, start = 1947, end = 1972)
lnwin2 <- window(lngdp, start = 1972, end = 1997)
lnwin3 <- window(lngdp, start = 1997, end = 2019)

par(mfrow = c(1, 3))
plot(lnwin1, ylim = c(min(lngdp), max(lngdp)))
plot(lnwin2, ylim = c(min(lngdp), max(lngdp)))
plot(lnwin3, ylim = c(min(lngdp), max(lngdp)))
```

```{r}
par(mfrow = c(1, 3))
boxplot(lnwin1, ylim = c(min(lngdp), max(lngdp)))
boxplot(lnwin2, ylim = c(min(lngdp), max(lngdp)))
boxplot(lnwin3, ylim = c(min(lngdp), max(lngdp)))
```

The Mean-Variance plot reinforces this.

*Note: In this case, `ncol = 36` is chosen because it is the number that includes around 8 observations per point.*

```{r}
mean = apply(matrix(lngdp, ncol = 36), 2, mean)
var = apply(matrix(lngdp, ncol = 36), 2, var)
plot(var ~ mean, main = "Mean-Variance plot", ylim = c(0, 0.15))
abline(lm(var ~ mean))
```

### Is seasonality present?

No, as the time series has already been seasonally adjusted.

### Is the mean constant?

We can assess whether the mean is constant or not by computing the ACF and the PACF.

```{r}
par(mfrow = c(1, 2))
acf(lngdp, ylim = c(-1, 1), lag.max = 60, lwd = 2)
pacf(lngdp, ylim = c(-1, 1), lag.max = 60, lwd = 2)
```

As we can observe, the mean is not constant, which goes in line with the fact that the time series presents an overall upward trend.

Nevertheless, since such trend is linear, it is possible to get rid of it by differentiating.

```{r}
par(mfrow = c(1, 1))
d1lngdp <- diff(lngdp)
plot(d1lngdp)
abline(h = 0)
abline(h = mean(d1lngdp), col = 2)
```

As we can observe, the plot has changed drastically, and now constant mean (different from 0) is noticeable.

Moreover, comparing variances allows us to analyze if extra differences are needed.

```{r}
formatC(var(lngdp), format = "f", digits = 7)
formatC(var(d1lngdp), format = "f", digits = 7)
formatC(var(diff(d1lngdp)), format = "f", digits = 7)
```

The variance decreases substantially after computing the first difference, but it barely moves from the first to the second. Thus, no extra differences should be added.

Let us now plot the updated ACF and PACF.

```{r}
par(mfrow = c(1, 2))
acf(d1lngdp, ylim = c(-1, 1), lag.max = 60, lwd = 2)
pacf(d1lngdp, ylim = c(-1, 1), lag.max = 60, lwd = 2)
```

The ACF swiftly decays to 0, meaning that only the very first lags are significant.

**The transformed time series is stationary.**

```{r}
par(mfrow = c(1, 2))
plot(gdpusa, main = "X_t")
plot(d1lngdp, main = "W_t")
abline(h = 0)
```


## GDPUSAvar

Firstly, we import the data.

```{r}
gdpusavar <- ts(read.table("gdpusavar.dat"), start = 1990, freq = 4)
gdpusavar
```

We plot the time series so as to visualize its shape.

```{r}
plot(gdpusavar, xlab = "Year", ylab = "US GDP (quarterly change in %)", type = "o")
abline(h = mean(gdpusavar), col = 2)
```

In this case, since data is on the quarterly rate of change of US GDP, we observe a time series that fluctuates between positive and negative values, with the biggest outliers happening in 2008, coinciding with that year's financial crisis.

### Is the variance constant?

*Note: Grouping data every two years caused issues with some plots, so we opted to keep working with the time series as it is.*

Let us compute the boxplot for each year in the time series.

```{r}
boxplot(gdpusavar ~ floor(time(gdpusavar)))
```

```{r}
mean = apply(matrix(gdpusavar, ncol = 15), 2, mean)
var = apply(matrix(gdpusavar, ncol = 15), 2, var)
plot(var ~ mean, main = "Mean-Variance plot")
```

### Is seasonality present?

No, as the time series has already been seasonally adjusted.

### Is the mean constant?

```{r}
plot(gdpusavar, xlab = "Year", ylab = "US GDP (quarterly change in %)")
abline(h = mean(gdpusavar), col = 2)
```

```{r}
formatC(var(gdpusavar), format = "f", digits = 7)
formatC(var(diff(gdpusavar)), format = "f", digits = 7)
```

```{r}
par(mfrow = c(1, 2))
acf(gdpusavar, ylim = c(-1, 1), lag.max = 60, lwd = 2)
pacf(gdpusavar, ylim = c(-1, 1), lag.max = 60, lwd = 2)
```