---
title: 'Financial Econometrics Practice TS #2 (Forecasting)'
subtitle: Daniel Campos, Joshua Castillo, & Cherryl Chico; February 22, 2026
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#set working directory to the folder where this file is located
setwd("/Users/cherrylchico/Desktop/BSE/T2/Financial Econometrics/finecon/hw2")
```
# Load Packages and Data
```{r}
library(forecast)
library(tseries)
library(lmtest)
library(nortest)
library(MASS)
```
Read the data as a time series object. The data is the number of individuals registered as unemployed in SEPE oﬃces since January 2000 available at https://expinterweb.mites.gob.es/series/. Select window starting 2000.

```{r}
series <- window(
  ts(read.table("unemploy.dat", dec = ",") / 1e6,
     start = 1996, freq = 12),
  start = c(2000),
  end = c(2025, 12)
)

# Display properties
cat("Class:", class(series), "\n")
cat("Length:", length(series), "observations\n")
cat("Start:", start(series)[1], "Period", start(series)[2], "\n")
cat("End:", end(series)[1], "Period", end(series)[2], "\n")
cat("Frequency:", frequency(series), "\n\n")

# show values
round(series, 3)
```


```{r}
# constants
series_name <- "Unemployment Rate"
start_year <- 2000
start_period <- 1
freq <- 12  
period_length <- 12 
xlabel <- "month"
ylabel <- "unemployment rate"
```
```
some obvious yearly seasonality, but SARIMA will handle that
```{r}
plot(series, main=series_name, xlab=xlabel, ylab=ylabel)
abline(v=1996:2026,col=4,lty=3)
```
To model, we first achieve variance constancy and let SARIMA handle the seasonality and mean stability. 

# Transformation to Variance Constancy
use rolling window stdev better than boxplot, with heteroskedasticity
```{r}
#function to plot rolling standard deviation
plot_rolling_std <- function(series, window_size, name="Series") {
  rolling_std <- sapply(1:(length(series) - window_size + 1), function(i) {
    sd(series[i:(i + window_size - 1)])
  })
  
  plot(rolling_std, type = "l",
    main = paste(name, "\nRolling Standard Deviation (",
           window_size, "-month window)", sep = ""),
    xlab = "Time", ylab = "Standard Deviation")
  abline(h=mean(rolling_std), col=2, lwd=2)
}

```

## Rolling standard deviation
```{r rolling_std}
plot_rolling_std(series, 12, series_name)  
```
## LN Transformation
seemed to handle mean stability and seasonality
```{r log}
ln_transformed <- log(series / 100)
plot(ln_transformed, type="o", 
  main=paste(series_name, "(log transformed)"),
  ylab=paste("Log of Unemployment Rate (%)", sep=""),
  xlab="")
abline(h=mean(ln_transformed), col=2, lwd=2)
```

still with obvious heteroskedasticity here
```{r rolling std returns}
plot_rolling_std(ln_transformed, 12, paste(series_name, "(log transformed)"))
```
##  Box Cox Transformation
```{r boxcox}
# Box-Cox transformation to find optimal lambda for variance stabilization
boxcox_result <- boxcox(series ~ 1, lambda = seq(-2, 2, by = 0.1))
optimal_lambda <- boxcox_result$x[which.max(boxcox_result$y)]
cat("Optimal lambda for Box-Cox transformation:", optimal_lambda, "\n")
# Apply Box-Cox transformation with optimal lambda
bc_transformed <- (series^optimal_lambda - 1) / optimal_lambda
```
```{r}
plot(bc_transformed, type="o", 
  main=paste(series_name, "(Box-Cox transformed)"),
  ylab=paste("Box-Cox transformed Unemployment Rate (%)", sep=""),
  xlab="")
abline(h=mean(bc_transformed), col=2, lwd=2)
```
plot
```{r}
plot_rolling_std(bc_transformed, 12, paste(series_name, "(Box-Cox transformed)"))
```

plot all three transformations together for comparison
```{r}
par(mfrow=c(3,1))
plot_rolling_std(series, 12, paste(series_name, "(Original)"))
plot_rolling_std(bc_transformed, 12, paste(series_name, "(Box-Cox transformed)"))
plot_rolling_std(ln_transformed, 12, paste(series_name, "(log transformed)"))
```
# Prep for SARIMA
```{r}
par(mfrow=c(1,1))
working_series <- ln_transformed
working_series_name <- paste(series_name, "(log transformed)")
ylabel <- paste("Log of", series_name, "(%)", sep="")
plot(working_series, type="o", main=working_series_name, ylab=ylabel, xlab=xlabel)
```
ACF PACF to see potential sarima models

definitely need differencing
```{r}
par(mfrow=c(1,2))
acf(working_series, ylim=c(-1,1), lag.max=max(60, 3*freq), 
    lwd=2, main=paste("ACF:", working_series_name),
)
pacf(working_series, ylim=c(-1,1), lag.max=max(60, 3*freq), 
     lwd=2, main=paste("PACF:", working_series_name),
     col=c(2, rep(1, 59)))
par(mfrow=c(1,1))
```
Do first differencing to determine seasonality and ARMA structure
```{r}
diff_series <- diff(working_series)
par(mfrow=c(1,2))
lag_limit <- max(60, 3*freq)
seasonal_lags <- seq(freq, lag_limit, by=freq)

acf_obj <- acf(diff_series, ylim=c(-1,1), lag.max=lag_limit,
         lwd=2, col=1, main=paste("ACF:", working_series_name, "(1st Diff)"))
acf_lags <- as.numeric(acf_obj$lag)[-1]
acf_vals <- as.numeric(acf_obj$acf)[-1]
acf_idx <- which(round(acf_lags * freq) %in% seasonal_lags)
segments(acf_lags[acf_idx], 0, acf_lags[acf_idx], acf_vals[acf_idx], col=2, lwd=2)
points(acf_lags[acf_idx], acf_vals[acf_idx], col=2, pch=16)

pacf_obj <- pacf(diff_series, ylim=c(-1,1), lag.max=lag_limit,
         lwd=2, col=1, main=paste("PACF:", working_series_name, "(1st Diff)"))
pacf_lags <- as.numeric(pacf_obj$lag)
pacf_vals <- as.numeric(pacf_obj$acf)
pacf_idx <- which(round(pacf_lags * freq) %in% seasonal_lags)
segments(pacf_lags[pacf_idx], 0, pacf_lags[pacf_idx], pacf_vals[pacf_idx], col=2, lwd=2)
points(pacf_lags[pacf_idx], pacf_vals[pacf_idx], col=2, pch=16)
```
We should do seasonality 12 for both AR and MA, and then up to AR(5) and MA(3) for the non-seasonal part, based on the ACF and PACF patterns.


```{r}
working_series <- diff_series
```
# SARIMA Modeling

## SARIMA function
```{r}
analyze_sarima <- function(series, order, seasonal) {
  
  # --- Setup ---
  s <- frequency(series)
  n <- length(series)
  model_name <- paste0("ARIMA(", paste(order, collapse=","), ")(", 
                       paste(seasonal$order, collapse=","), ")", s)
  
  cat("\n==================================================================\n")
  cat("  ANALYSIS FOR MODEL:", model_name, "\n")
  cat("==================================================================\n")
  
  # -----------------------------------------------------------------------
  # 1. Estimation & Significance Check
  # -----------------------------------------------------------------------
  fit_full <- arima(series, order = order, seasonal = seasonal)
  
  diag_var <- diag(fit_full$var.coef)
  if (any(diag_var <= 0)) {
    cat("\n⚠️  CRITICAL WARNING: The Hessian matrix is unstable (Negative Variances).\n")
    se_vec <- ifelse(diag_var > 0, sqrt(diag_var), NA)
  } else {
    se_vec <- sqrt(diag_var)
  }
  
  t_vec  <- fit_full$coef / se_vec
  p_vec  <- 2 * (1 - pnorm(abs(t_vec)))
  
  coeftest <- data.frame(
    Estimate = fit_full$coef,
    Std.Error = se_vec,
    t_stat = t_vec,
    p_value = p_vec
  )
  
  coeftest$Significant <- ifelse(!is.na(coeftest$p_value) & 
                                   (coeftest$p_value < 0.05 | abs(coeftest$t_stat) > 2), 
                                 "Yes", "No")
  
  cat("\n[1] COEFFICIENT SIGNIFICANCE (Alpha = 0.05 or |t| > 2):\n")
  print(round(coeftest[,1:4], 4))
  cat("\nSignificance Flags:\n")
  print(coeftest[,5, drop=FALSE])
  
  # -----------------------------------------------------------------------
  # 2. Residual Analysis (Diagnostics)
  # -----------------------------------------------------------------------
  resid <- residuals(fit_full)
  
  # A. Causality / Invertibility Roots (Min Modulus)
  # AR Roots
  ar_poly <- fit_full$model$phi
  if (length(ar_poly) > 0 && any(ar_poly != 0)) {
    ar_roots_mod <- Mod(polyroot(c(1, -ar_poly)))
    min_ar_root  <- min(ar_roots_mod)
  } else {
    min_ar_root  <- NA
  }
  
  # MA Roots
  ma_poly <- fit_full$model$theta
  if (length(ma_poly) > 0 && any(ma_poly != 0)) {
    ma_roots_mod <- Mod(polyroot(c(1, ma_poly)))
    min_ma_root  <- min(ma_roots_mod)
  } else {
    min_ma_root  <- NA
  }
  
  is_causal     <- ifelse(!is.na(min_ar_root), min_ar_root > 1, TRUE)
  is_invertible <- ifelse(!is.na(min_ma_root), min_ma_root > 1, TRUE)
  
  cat("\n[2] DIAGNOSTICS:\n")
  cat("Min AR Root Modulus: ", min_ar_root, "\n")
  cat("Min MA Root Modulus: ", min_ma_root, "\n")
  
  # B. Plots
  par(mfrow=c(2,2), mar=c(4,4,3,2))
  plot(resid, main="Residuals", ylab="Error")
  abline(h=0, col="gray"); abline(h=c(-3*sd(resid), 3*sd(resid)), lty=2, col="red")
  scatter.smooth(sqrt(abs(resid)), main="Variance Check", ylab="sqrt(|Resid|)", lpars=list(col="red"))
  qqnorm(resid); qqline(resid, col="red")
  hist(resid, breaks=20, freq=FALSE, main="Histogram"); curve(dnorm(x, mean=mean(resid), sd=sd(resid)), col="red", add=TRUE)
  
  # C. Tests
  sw_test <- shapiro.test(resid)
  lb_test <- Box.test(resid, lag = 20, type = "Ljung-Box")
  
  # -----------------------------------------------------------------------
  # 3. Forecasting & Cross-Validation
  # -----------------------------------------------------------------------
  train_end <- c(floor(time(series)[n-12]), cycle(series)[n-12])
  train_set <- window(series, end = train_end)
  test_set  <- window(series, start = time(series)[n-11]) 
  
  fit_train <- arima(train_set, order = order, seasonal = seasonal)
  fc <- predict(fit_train, n.ahead = 12)
  
  fc_orig <- exp(fc$pred); test_orig <- exp(test_set)
  lower_bound <- exp(fc$pred - 1.96 * fc$se); upper_bound <- exp(fc$pred + 1.96 * fc$se)
  
  # -----------------------------------------------------------------------
  # 4. Summary Table Generation
  # -----------------------------------------------------------------------
  summary_row <- data.frame(
    Model_Name      = model_name,
    AIC             = fit_full$aic,
    BIC             = BIC(fit_full),
    Sigma2          = fit_full$sigma2,
    
    Resid_Norm_PVal = sw_test$p.value,
    Normal       = ifelse(sw_test$p.value > 0.05, "Yes", "No"),
    
    Resid_Ind_PVal  = lb_test$p.value,
    Independent  = ifelse(lb_test$p.value > 0.05, "Yes", "No"),
    
    # Positioned as requested: Root value to the left of the Boolean status
    Min_AR_Root     = min_ar_root,
    Causal          = ifelse(is_causal, "Yes", "No"),
    
    Min_MA_Root     = min_ma_root,
    Invertible      = ifelse(is_invertible, "Yes", "No"),
    
    RMSE            = sqrt(mean((test_orig - fc_orig)^2)),
    RMSPE           = sqrt(mean(((test_orig - fc_orig) / test_orig)^2)) * 100,
    MAE             = mean(abs(test_orig - fc_orig)),
    MAPE_Pct        = mean(abs((test_orig - fc_orig) / test_orig)) * 100,
    Mean_CI_Width   = mean(upper_bound - lower_bound)
  )
  
  cat("\n[3] PERFORMANCE SUMMARY:\n")
  print(t(summary_row))
  
  return(summary_row)
}
```
## Comparison Function
```{r}
# -- COMPARATIVE ANALYSIS FUNCTION --


compare_sarima_models <- function(series, models_list) {
  
  cat("\n--- Starting Automated Model Comparison ---\n")
  
  # This will store the summary row for each successful model
  comparison_results <- list()
  
  # Loop through each model specification in the list
  for (i in seq_along(models_list)) {
    spec <- models_list[[i]]
    
    cat(paste0("\nProcessing Model ", i, " of ", length(models_list), "...\n"))
    
    # Use tryCatch to prevent the loop from stopping if a model fails to converge
    result <- tryCatch({
      analyze_sarima(series = series, 
                     order = spec$order, 
                     seasonal = spec$seasonal)
    }, error = function(e) {
      cat(paste0("⚠️ Error in Model ", i, ": ", e$message, "\n"))
      return(NULL) # Return NULL so we can skip it
    })
    
    # If the model was successful, add it to our list
    if (!is.null(result)) {
      comparison_results[[length(comparison_results) + 1]] <- result
    }
  }
  
  # Combine all the rows into one master table
  final_table <- do.call(rbind, comparison_results)
  
  # Sort by BIC by default (it penalizes complexity more than AIC)
  final_table <- final_table[order(final_table$BIC), ]
  
  cat("\n--- Comparison Complete ---\n")
  return(final_table)
}
```
## Models
```{r}
# All possible non-seasonal ARMA(p,q) with p <= 5 and q <= 3
# Seasonal period fixed at 12
model_grid <- expand.grid(p = 1:5, q = 1:3)

my_models <- lapply(seq_len(nrow(model_grid)), function(i) {
  list(
    order = c(model_grid$p[i], 0, model_grid$q[i]),
    seasonal = list(order = c(1,0,1), period = 12)
  )
})

# Print all model names
model_names <- paste0(
  "ARIMA(", model_grid$p, ",0,", model_grid$q, ")",
  "(1,0,1)[12]"
)
print(model_names)
```
show a few of the models to confirm they are correct
```{r}
head(my_models, 1)
```

## Run all models
```{r}
# Run Analysis
comparison_table <- compare_sarima_models(working_series, my_models)
print(comparison_table)
```

# Choosing Model
All modeels dont have normal residuals because of covid. Disregarding that, based on model performance we have models 13, 3, 6 but for parsimony were choosing the simplest which is ARIMA(1,0,2)(1,0,1)[12].

```{r}
analyze_sarima(working_series, order = c(1,0,2), seasonal = list(order = c(1,0,1), period = 12))
```
